---
title: "Calculations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{calculations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(OCepi)
library(dplyr)
```

## add_percent()

One of the most used calculations is proportion. If you do not want to multiply proportion by 100, add argument `multiply = FALSE`.

```{r prop}
linelist |>
  mutate(race_ethnicity = recode_race(Ethnicity, Race, abbr_names = TRUE)) |>
  count(race_ethnicity) |>
  mutate(percent = add_percent(n, digits = 1))
```

When preparing data for {ggplot2}, you may want to create labels with n and percent. Use argument `n_suppress` to suppress low values and `reverse` to set order of n and % (default is n then %).

```{r prop2}
linelist |>
  mutate(race_ethnicity = recode_race(Ethnicity, Race, abbr_names = TRUE)) |>
  count(race_ethnicity) |>
  mutate(
    percent = add_percent(n, digits = 1),
    label = n_percent(n, percent, reverse = TRUE, n_suppress = 10)
    )
```

## rate_per_100k()

Another common calculation for our team are is incidence rate per 100,000.

```{r rate}
linelist |>
  mutate(race_ethnicity = recode_race(Ethnicity, Race, abbr_names = TRUE)) |>
  filter(race_ethnicity %in% c("White","NHOPI","Multiple Races")) |>
  count(race_ethnicity) |>
  mutate(incidence_rate = case_when(
    race_ethnicity == "White" ~ rate_per_100k(n, 1500000, digits = 1),
    race_ethnicity == "NHOPI" ~ rate_per_100k(n, 100000, digits = 1),
    race_ethnicity == "Multiple Races" ~ rate_per_100k(n, 200000, digits = 1)
    )
  )
```

## time_between()

`time_between()` calculates the time elapsed between two dates in the following units: days, weeks, months and years. Value is always rounded down to account for complete elapsed time. Note: order of dates does not necessarily matter, but beware of sign.

```{r time-between}
episode_date = as.Date("2020-04-01")
death_date = as.Date("2020-05-15")

time_between(death_date, episode_date, unit = "days")
time_between(episode_date, death_date, unit = "days")

dob = as.Date("1986-04-01")
vaccine_date = as.Date("2023-05-15")

time_between(vaccine_date, dob, unit = "years")
time_between(dob, vaccine_date, unit = "years")
```
