---
title: "Data Cleaning"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{data-cleaning}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message=FALSE, warning=FALSE, echo=FALSE}
library(OCepi)
library(dplyr)
library(ggplot2)
linelist <- "Ethnicity,Race,Gender,Age,SexualOrientation,SpecimenDate
Non-Hispanic or Latino,Multiple Races,M,46,HET,6/7/2022
Unknown,Unknown,M,4,HET,6/9/2022
Non-Hispanic or Latino,White,F,52,UNK,6/7/2022
Non-Hispanic or Latino,White,F,77,UNK,6/11/2022
Unknown,American Indian or Alaska Native,M,71,HET,6/10/2022
Non-Hispanic or Latino,Other,M,70,HET,6/9/2022
Non-Hispanic or Latino,Black or African American,F,11,HET,6/8/2022
Non-Hispanic or Latino,Black or African American,F,8,HET,6/12/2022
Hispanic or Latino,American Indian or Alaska Native,F,41,HET,6/12/2022
Non-Hispanic or Latino,Black or African American,M,56,HET,6/10/2022
Non-Hispanic or Latino,American Indian or Alaska Native,F,40,HET,6/14/2022
Non-Hispanic or Latino,Black or African American,M,1,HET,6/8/2022
Non-Hispanic or Latino,Other,F,60,HET,6/7/2022
Hispanic or Latino,American Indian or Alaska Native,F,22,HET,6/10/2022
Hispanic or Latino,Native Hawaiian or Other Pacific Islander,M,84,HET,6/6/2022
Non-Hispanic or Latino,Unknown,F,16,HET,6/13/2022
Hispanic or Latino,White,M,46,HET,6/7/2022
Non-Hispanic or Latino,Unknown,U,6,HET,6/13/2022
Hispanic or Latino,Native Hawaiian or Other Pacific Islander,F,73,HET,6/8/2022
Non-Hispanic or Latino,Black or African American,F,81,HET,6/8/2022
Non-Hispanic or Latino,Black or African American,M,51,DNK,6/4/2022
Non-Hispanic or Latino,Native Hawaiian or Other Pacific Islander,M,2,HET,6/11/2022
Non-Hispanic or Latino,Other,M,18,HET,6/7/2022
Hispanic or Latino,Native Hawaiian or Other Pacific Islander,F,25,HET,6/15/2022
Non-Hispanic or Latino,White,U,7,HET,6/7/2022
Non-Hispanic or Latino,Other,F,35,HET,6/7/2022
Non-Hispanic or Latino,Other,F,67,HET,6/10/2022
Non-Hispanic or Latino,Native Hawaiian or Other Pacific Islander,F,60,HET,6/10/2022
Hispanic or Latino,Asian,U,62,UNK,6/6/2022
Non-Hispanic or Latino,Other,F,61,HET,6/4/2022
Non-Hispanic or Latino,Other,F,71,HET,6/5/2022
Non-Hispanic or Latino,White,F,5,HET,6/7/2022
Non-Hispanic or Latino,American Indian or Alaska Native,F,27,HET,6/11/2022
Unknown,Native Hawaiian or Other Pacific Islander,M,13,HET,6/7/2022
Non-Hispanic or Latino,Native Hawaiian or Other Pacific Islander,M,42,HET,6/13/2022
Hispanic or Latino,White,U,34,HET,6/11/2022
Non-Hispanic or Latino,White,M,32,HET,6/12/2022
Non-Hispanic or Latino,Black or African American,F,9,HET,6/4/2022
Non-Hispanic or Latino,White,D,55,HET,6/9/2022
Non-Hispanic or Latino,Black or African American,M,33,HET,6/10/2022
Non-Hispanic or Latino,American Indian or Alaska Native,U,87,UNK,6/10/2022
Non-Hispanic or Latino,White,F,86,DNK,6/10/2022
Non-Hispanic or Latino,Asian,F,20,HET,6/9/2022
Non-Hispanic or Latino,White,U,24,DNK,6/8/2022
Hispanic or Latino,American Indian or Alaska Native,U,45,UNK,6/11/2022
Non-Hispanic or Latino,Native Hawaiian or Other Pacific Islander,M,58,HET,6/13/2022
Hispanic or Latino,White,F,89,HET,6/13/2022
Hispanic or Latino,Asian,M,48,HET,6/11/2022
Non-Hispanic or Latino,Other,M,42,HET,6/13/2022
Non-Hispanic or Latino,White,F,56,HET,6/9/2022
Non-Hispanic or Latino,Native Hawaiian or Other Pacific Islander,F,1,HET,6/9/2022
Non-Hispanic or Latino,Black or African American,M,25,HET,6/7/2022
Non-Hispanic or Latino,Asian,F,58,UNK,6/9/2022
Non-Hispanic or Latino,White,F,39,DNK,6/7/2022
Non-Hispanic or Latino,Asian,M,38,HET,6/14/2022
Non-Hispanic or Latino,Multiple Races,M,58,HET,6/12/2022
Non-Hispanic or Latino,Black or African American,F,12,HET,6/15/2022
Non-Hispanic or Latino,American Indian or Alaska Native,M,40,HET,6/8/2022
Non-Hispanic or Latino,Native Hawaiian or Other Pacific Islander,F,52,HET,6/9/2022
Non-Hispanic or Latino,Multiple Races,F,10,HET,6/12/2022
Non-Hispanic or Latino,Black or African American,M,29,HET,6/10/2022
Non-Hispanic or Latino,Native Hawaiian or Other Pacific Islander,F,15,HET,6/9/2022
Non-Hispanic or Latino,American Indian or Alaska Native,F,22,HET,6/5/2022
Hispanic or Latino,American Indian or Alaska Native,F,75,HET,6/11/2022
Unknown,White,F,85,HET,6/12/2022
Non-Hispanic or Latino,American Indian or Alaska Native,F,72,HET,6/16/2022
Non-Hispanic or Latino,American Indian or Alaska Native,U,14,HET,6/18/2022
Non-Hispanic or Latino,Asian,F,80,HET,6/12/2022
Hispanic or Latino,Asian,F,32,HET,6/13/2022
Non-Hispanic or Latino,Asian,M,75,HET,6/10/2022
Non-Hispanic or Latino,Unknown,D,71,HOM,6/11/2022
Non-Hispanic or Latino,Black or African American,D,17,HET,6/9/2022
Non-Hispanic or Latino,Other,F,46,HET,6/7/2022
Non-Hispanic or Latino,Other,M,83,UNK,6/9/2022
Non-Hispanic or Latino,Asian,M,79,HET,6/8/2022
Non-Hispanic or Latino,Unknown,M,20,HET,6/11/2022
Hispanic or Latino,Native Hawaiian or Other Pacific Islander,M,61,HET,6/9/2022
Non-Hispanic or Latino,Black or African American,M,32,HET,6/10/2022
Non-Hispanic or Latino,White,F,50,HET,6/14/2022
Hispanic or Latino,Unknown,D,85,HOM,6/13/2022
Non-Hispanic or Latino,Native Hawaiian or Other Pacific Islander,F,4,HET,6/9/2022
Non-Hispanic or Latino,Native Hawaiian or Other Pacific Islander,F,59,HET,6/8/2022
Unknown,Black or African American,M,75,HET,6/9/2022
Non-Hispanic or Latino,Multiple Races,M,45,HET,6/13/2022
Non-Hispanic or Latino,Black or African American,U,56,HET,6/9/2022
Non-Hispanic or Latino,Other,F,36,HET,6/11/2022
Non-Hispanic or Latino,American Indian or Alaska Native,M,79,HET,6/6/2022
Hispanic or Latino,American Indian or Alaska Native,M,88,HET,6/9/2022
Hispanic or Latino,Other,F,82,HET,6/16/2022
Non-Hispanic or Latino,Black or African American,F,7,HET,6/8/2022
Non-Hispanic or Latino,Unknown,D,26,HET,6/11/2022
Non-Hispanic or Latino,Multiple Races,F,75,DNK,6/11/2022
Unknown,Other,D,8,HET,6/8/2022
Non-Hispanic or Latino,Unknown,M,77,HET,6/15/2022
Non-Hispanic or Latino,American Indian or Alaska Native,M,14,HET,6/14/2022
Non-Hispanic or Latino,Other,F,71,HOM,6/8/2022
Non-Hispanic or Latino,Other,F,23,HET,6/10/2022
Non-Hispanic or Latino,White,F,16,HET,6/6/2022
Non-Hispanic or Latino,Asian,F,43,BIS,6/9/2022
Hispanic or Latino,Native Hawaiian or Other Pacific Islander,M,74,HET,6/8/2022
Unknown,White,M,70,HET,6/14/2022
Non-Hispanic or Latino,Multiple Races,M,39,HET,6/10/2022
Non-Hispanic or Latino,Unknown,F,31,HET,6/14/2022
Hispanic or Latino,Asian,M,38,HET,6/9/2022
Non-Hispanic or Latino,Other,F,31,HET,6/8/2022"

header.true <- function(df){
  names(df) <- as.character(unlist(df[1,]))
  df[-1,]
}

linelist <- dplyr::as_tibble(linelist) %>%
  tidyr::separate_rows(value, sep = "\n") %>%
  tidyr::separate_wider_delim(value, delim = ",", names = paste0("V",1:6))
linelist <- header.true(linelist)
linelist$Age <- as.numeric(linelist$Age)
linelist$SpecimenDate <- as.Date(linelist$SpecimenDate, "%m/%d/%Y")
```

The functions in this package were designed to simplify and standardize the most frequent recoding tasks. The examples below will use simulated outbreak data:

```{r glance, message=FALSE, warning=FALSE, echo=FALSE}
head(linelist, 10)
```

# Recoding

## Core Functions:

-   `recode_race` - able to accept: ethnicity and race, race only, LOINC codes, and 'Multi-race Status' variable from VRBIS.

    -   Set `abbr_names` to TRUE to abbreviate long category names i.e. "American Indian or Alaska Native" to "AI/AN".

    -   If using ethnicity and race, use order `recode_race(ethnicity, race)`.

-   `recode_gender` - can be used with CalREDIE or VRBIS; aims to use the most inclusive terms possible.

-   `recode_orientation` - created specifically for CalREDIE. Expects CTCIAdtlDemOrient variable.

-   `age_groups` - has several presets for grouping age:

    -   **cair2 peds**: \<12 M, 12-15 M, 16-23 M, 24 M, 25-47 M, 4-6 Years, 6+ Years
    -   **covid:** 0-17, 18-24, 25-34, 35-44, 45-54, 55-64, 65-74, 75-84, 85+
    -   **decade:** 0-9, 10-19, 20-29, 30-39, 40-49, 50-59, 60-69, 70-79, 80+
    -   **enteric:** 0-4, 5-14, 15-24, 25-44, 45-64, 65+
    -   **flu vax:** 0-18, 19-49, 50-64, 65+
    -   **hcv:** 0-17, 18-39, 40-59, 60+
    -   **mpox:** 0-15, 16-24, 25-34, 35-44, 45-54, 55-64, 65+
    -   **school:** 0-4, 5-11, 12-17, 18-64, 65+
    -   **wnv:** 0-17, 18-24, 25-34, 35-44, 45-54, 55-64, 65+

Example:

```{r demographics}
linelist <- linelist %>%
  mutate(
    Gender = recode_gender(Gender),
    race_ethnicity = recode_race(Ethnicity, Race, abbr_names = TRUE),
    age_group = age_groups(Age, type = "school"),
    SexualOrientation = recode_orientation(SexualOrientation)
  )
```

Frequency table with percentage using `add_percent` and incidence rates using `rate_per_100k`:

```{r freq, message=FALSE, warning=FALSE}
linelist %>%
  count(age_group) %>%
  mutate(
    percent = add_percent(n),
    rate = case_when(
      age_group == "0-4" ~ rate_per_100k(n, 100000, digits = 1),
      age_group == "5-11" ~ rate_per_100k(n, 300000, digits = 1),
      age_group == "12-17" ~ rate_per_100k(n, 500000, digits = 1),
      age_group == "18-64" ~ rate_per_100k(n, 1000000, digits = 1),
      age_group == "65+" ~ rate_per_100k(n, 2000000, digits = 1)
      )
    )
```

## Addresses

Our best attempt to standardize and clean addresses is through `clean_address`. This function will: convert characters to title casing, recode abbreviated cardinal directions, and recode abbreviated street endings (i.e. Street, Circle, Plaza). On occasion there is a need to drop additional address data (i.e. Apartment, Unit), which can be done using `keep_extra`.

```{r clean-address}
cases <- data.frame(
  Address = c("1234 Main Street Apt 204","501 N Capital St","233 W Green Plz Unit 3")
)

cases %>%
  mutate(
    Address_clean = clean_address(Address, keep_extra = TRUE)
    )
```

## Phone Numbers

We also attempt to standardize phone number to a 10 digit U.S. format. If after removing country code and symbols, and length is not 10 characters, returned value is NA.

```{r phone}
cases <- data.frame(
  HomePhone = c("1-714-777-1234","(949) 555-1234","+442071539000")
)

cases %>%
  mutate(HomePhone_clean = clean_phone(HomePhone))
```

## Census Tracts

Census tract across datasets are not always the same - some contain state and county fips codes, some do not. To make them match, use `recode_ctract` to remove state and county fips code.

```{r extra-recode}
recode_ctract("06059099244")
```

## VRBIS

Core functions:

-   `vrbis_manner_death` - expects "Manner of Death" variable, follows CCDF data dictionary
-   `vrbis_place_death` - expects "Place of Death (Facility)", follows CCDF data dictionary
-   `vrbis_resident` - determine if death belongs to your local health jurisdiction

```{r vrbis1, message = FALSE, warning = FALSE}
vrbis_manner_death("A")

vrbis_place_death(6)
```

To use `vrbis_resident`, you will need to know your county fips code and CCDF county code (from appendix G). Returned output is 0/1 where 1 indicates death belongs to your county. For columns, you'll need (in order): "Place of Death (Facility)", "County of Death (Code)", "Decedents County of Residence (NCHS Code)". Next, plug in your fips code to `fips` and CCDF county code to `county` arguments.

```{r vrbis2, message = FALSE, warning = FALSE, eval = FALSE}
df <- df %>%
  mutate(
    lhj_resident = vrbis_resident(
      place_of_death,
      county_of_death_code,
      decedents_county_of_residence_NCHS_Code,
      fips = "059",
      county = "30"
      )
    ) %>%
  filter(lhj_resident == 1)
```

# Data Masking

## Suppression

Small cell sizes may need to be suppressed to protect patient confidentiality prior to reporting. In this example, any cell sizes `less_than` 10 will be suppressed and `replace_with` double asterisk.

```{r suppress}
linelist %>% count(age_group) %>% mutate(n_suppress = suppress(n, less_than = 10, replace_with = "**"))
```

## Redaction

Depending on unit/program policy, any mention of HIV or AIDS may need to be removed from the dataset. While not yet exhaustive, this function accounts for several variations of HIV and AIDS. If found, a warning message is printed to the console.

```{r redact}
df <- data.frame(cause = c("cancer","hepatitis","COVID-19","HIV"))
print(df)

df <- hiv_redact(df)
print(df)
```

# CAIR2

## Baby Name

Removes variants of baby/twin/newborn from first name. Addresses data quality issue in birth hepatitis B doses.

```{r, babyname}
baby_name("baby BOY")

baby_name("NEWBORN")

baby_name("twin A")
```

# Data Management

## Data.frames

`remove_empty_cols` drops all columns from data.frame that are blank (represented by `NA` or "")

```{r, message = TRUE, warning = TRUE}
df <- data.frame(a = c(NA,NA,NA), b = c("","",""), c = c(1,2,3))
print(df)

df <- remove_empty_cols(df)
print(df)
```
