---
title: "Date Conversions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{date-conversions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(OCepi)
library(dplyr)

linelist <- linelist
```

## MMWR

Core Functions:

-   `mmwr_year()` - calculate epidemiological year
-   `mmwr_week()` - calculate epidemiological week (or disease week)
-   `week_ending_date()` - calculate Saturday of disease week
-   `mmwr_calendar()` - returns data.frame with columns for disease week, week start and end date, for a given epidemiological year
-   `mmwrweek_to_date()` - calculate week ending date from epidemiological year and week

```{r mmwr-forward, message = FALSE, warning = FALSE}
dates <- linelist |>
  select(SpecimenDate) |>
  arrange(SpecimenDate) |>
  mutate(
    epi_year = mmwr_year(SpecimenDate),
    disease_week = mmwr_week(SpecimenDate),
    week_ending = week_ending_date(SpecimenDate)
  )

head(dates)
```

Convert epidemiological year and week to week ending date:

```{r mmwr-reverse}
dates <- dates |>
  select(epi_year, disease_week) |>
  mutate(
    week_ending = mmwrweek_to_date(epi_year, disease_week)
  )

head(dates)
```

To make `mmwr_calendar()` with total disease weeks + start/end dates:

```{r mmwr-cal, message = FALSE, warning = FALSE}
mmwr_calendar(2023) |>
  head(20)
```

## assign_season()

For epidemiologists working viral respiratory surveillance, patients/laboratory results can also be categorized by season. Season in this context spans week 40 of current year to week 39 of following year.

```{r resp-season, message = FALSE, warning = FALSE}
df <- data.frame(spec_date = as.Date(c("2023-10-01","2023-11-04","2024-09-28","2024-09-29")))

df |>
  mutate(season = assign_season(spec_date))
```

## to_month()

An alternative to grouping cases at the year or week level is by month. Returned output is a date formatted YYYY-MM-01.

```{r month, message = FALSE, warning = FALSE}
df <- data.frame(spec_date = as.Date(c("2023-10-01","2023-11-04","2024-09-28","2024-09-29")))

df |>
  mutate(month = to_month(spec_date))
```
