---
title: "date-conversions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{date-conversions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(OCepi)
library(dplyr)

linelist <- linelist
```

## MMWR Conversions

For most time series data, we prefer to aggregate at the week or year level for reporting purposes. Like most epidemiologists, we rely on MMWR disease week and year. For week, we have two flavors - `mwwr_week` and `week_ending_date` (Saturday of disease week). For year, we use `mmwr_year`.

```{r mmwr-forward, message = FALSE, warning = FALSE}
dates <- linelist |>
  select(SpecimenDate) |>
  arrange(SpecimenDate) |>
  mutate(
    epi_year = mmwr_year(SpecimenDate),
    disease_week = mmwr_week(SpecimenDate),
    week_ending = week_ending_date(SpecimenDate)
  )

head(dates)
```

## mmwrweek_to_date()

For those that need to convert epidemiological year and week to week ending date:

```{r mmwr-reverse}
dates <- dates |>
  select(epi_year, disease_week) |>
  mutate(
    week_ending = mmwrweek_to_date(epi_year, disease_week)
  )

head(dates)
```

## mmwr_calendar()

To make `mmwr_calendar()` with total disease weeks + start/end dates:

```{r mmwr-cal, message = FALSE, warning = FALSE}
mmwr_calendar(2023) |>
  head(20)
```

## assign_season()

For epidemiologists working viral respiratory surveillance, patients/laboratory results can also be categorized by season. Season in this context spans week 40 of current year to week 39 of following year.

```{r resp-season, message = FALSE, warning = FALSE}
df <- data.frame(spec_date = as.Date(c("2023-10-01","2023-11-04","2024-09-28","2024-09-29")))

df |>
  mutate(season = assign_season(spec_date))
```

## to_month()

An alternative to grouping cases at the year or week level is by month. Returned output is a date formatted YYYY-MM-01.

```{r month, message = FALSE, warning = FALSE}
df <- data.frame(spec_date = as.Date(c("2023-10-01","2023-11-04","2024-09-28","2024-09-29")))

df |>
  mutate(month = to_month(spec_date))
```
