---
title: "Extending ggplot2"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{extending-ggplot2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(OCepi)
library(dplyr)
library(ggplot2)

highlight <- data.frame(
  site = c("A","A","A","B","B","B"),
  locations = rep(letters[1:3],2),
  scores = c(89,62,74,88,89.5,82)
  )

cleveland <- data.frame(Country = c("UK-A","US-A","UK-B","US-B"), PRE = c(0,1,2,3), POST = c(4,5,6,7))

re <- data.frame(group = c("Native Hawaiian or Other Pacific Islander","Black or African American","American Indian or Alaska Native"), score = c(89.5, 84, 73))

year2023 <- seq.Date(from = as.Date("2023-01-01"), to = as.Date("2023-12-01"), by = "month")
ts <- data.frame(date = rep(year2023, 2), site = c(rep("A",12),rep("B",12)), scores = sample(50:95, 24))

linelist <- "Ethnicity,Race,Gender,Age,SexualOrientation,SpecimenDate
Non-Hispanic or Latino,Multiple Races,M,46,HET,6/7/2022
Unknown,Unknown,M,4,HET,6/9/2022
Non-Hispanic or Latino,White,F,52,UNK,6/7/2022
Non-Hispanic or Latino,White,F,77,UNK,6/11/2022
Unknown,American Indian or Alaska Native,M,71,HET,6/10/2022
Non-Hispanic or Latino,Other,M,70,HET,6/9/2022
Non-Hispanic or Latino,Black or African American,F,11,HET,6/8/2022
Non-Hispanic or Latino,Black or African American,F,8,HET,6/12/2022
Hispanic or Latino,American Indian or Alaska Native,F,41,HET,6/12/2022
Non-Hispanic or Latino,Black or African American,M,56,HET,6/10/2022
Non-Hispanic or Latino,American Indian or Alaska Native,F,40,HET,6/14/2022
Non-Hispanic or Latino,Black or African American,M,1,HET,6/8/2022
Non-Hispanic or Latino,Other,F,60,HET,6/7/2022
Hispanic or Latino,American Indian or Alaska Native,F,22,HET,6/10/2022
Hispanic or Latino,Native Hawaiian or Other Pacific Islander,M,84,HET,6/6/2022
Non-Hispanic or Latino,Unknown,F,16,HET,6/13/2022
Hispanic or Latino,White,M,46,HET,6/7/2022
Non-Hispanic or Latino,Unknown,U,6,HET,6/13/2022
Hispanic or Latino,Native Hawaiian or Other Pacific Islander,F,73,HET,6/8/2022
Non-Hispanic or Latino,Black or African American,F,81,HET,6/8/2022
Non-Hispanic or Latino,Black or African American,M,51,DNK,6/4/2022
Non-Hispanic or Latino,Native Hawaiian or Other Pacific Islander,M,2,HET,6/11/2022
Non-Hispanic or Latino,Other,M,18,HET,6/7/2022
Hispanic or Latino,Native Hawaiian or Other Pacific Islander,F,25,HET,6/15/2022
Non-Hispanic or Latino,White,U,7,HET,6/7/2022
Non-Hispanic or Latino,Other,F,35,HET,6/7/2022
Non-Hispanic or Latino,Other,F,67,HET,6/10/2022
Non-Hispanic or Latino,Native Hawaiian or Other Pacific Islander,F,60,HET,6/10/2022
Hispanic or Latino,Asian,U,62,UNK,6/6/2022
Non-Hispanic or Latino,Other,F,61,HET,6/4/2022
Non-Hispanic or Latino,Other,F,71,HET,6/5/2022
Non-Hispanic or Latino,White,F,5,HET,6/7/2022
Non-Hispanic or Latino,American Indian or Alaska Native,F,27,HET,6/11/2022
Unknown,Native Hawaiian or Other Pacific Islander,M,13,HET,6/7/2022
Non-Hispanic or Latino,Native Hawaiian or Other Pacific Islander,M,42,HET,6/13/2022
Hispanic or Latino,White,U,34,HET,6/11/2022
Non-Hispanic or Latino,White,M,32,HET,6/12/2022
Non-Hispanic or Latino,Black or African American,F,9,HET,6/4/2022
Non-Hispanic or Latino,White,D,55,HET,6/9/2022
Non-Hispanic or Latino,Black or African American,M,33,HET,6/10/2022
Non-Hispanic or Latino,American Indian or Alaska Native,U,87,UNK,6/10/2022
Non-Hispanic or Latino,White,F,86,DNK,6/10/2022
Non-Hispanic or Latino,Asian,F,20,HET,6/9/2022
Non-Hispanic or Latino,White,U,24,DNK,6/8/2022
Hispanic or Latino,American Indian or Alaska Native,U,45,UNK,6/11/2022
Non-Hispanic or Latino,Native Hawaiian or Other Pacific Islander,M,58,HET,6/13/2022
Hispanic or Latino,White,F,89,HET,6/13/2022
Hispanic or Latino,Asian,M,48,HET,6/11/2022
Non-Hispanic or Latino,Other,M,42,HET,6/13/2022
Non-Hispanic or Latino,White,F,56,HET,6/9/2022
Non-Hispanic or Latino,Native Hawaiian or Other Pacific Islander,F,1,HET,6/9/2022
Non-Hispanic or Latino,Black or African American,M,25,HET,6/7/2022
Non-Hispanic or Latino,Asian,F,58,UNK,6/9/2022
Non-Hispanic or Latino,White,F,39,DNK,6/7/2022
Non-Hispanic or Latino,Asian,M,38,HET,6/14/2022
Non-Hispanic or Latino,Multiple Races,M,58,HET,6/12/2022
Non-Hispanic or Latino,Black or African American,F,12,HET,6/15/2022
Non-Hispanic or Latino,American Indian or Alaska Native,M,40,HET,6/8/2022
Non-Hispanic or Latino,Native Hawaiian or Other Pacific Islander,F,52,HET,6/9/2022
Non-Hispanic or Latino,Multiple Races,F,10,HET,6/12/2022
Non-Hispanic or Latino,Black or African American,M,29,HET,6/10/2022
Non-Hispanic or Latino,Native Hawaiian or Other Pacific Islander,F,15,HET,6/9/2022
Non-Hispanic or Latino,American Indian or Alaska Native,F,22,HET,6/5/2022
Hispanic or Latino,American Indian or Alaska Native,F,75,HET,6/11/2022
Unknown,White,F,85,HET,6/12/2022
Non-Hispanic or Latino,American Indian or Alaska Native,F,72,HET,6/16/2022
Non-Hispanic or Latino,American Indian or Alaska Native,U,14,HET,6/18/2022
Non-Hispanic or Latino,Asian,F,80,HET,6/12/2022
Hispanic or Latino,Asian,F,32,HET,6/13/2022
Non-Hispanic or Latino,Asian,M,75,HET,6/10/2022
Non-Hispanic or Latino,Unknown,D,71,HOM,6/11/2022
Non-Hispanic or Latino,Black or African American,D,17,HET,6/9/2022
Non-Hispanic or Latino,Other,F,46,HET,6/7/2022
Non-Hispanic or Latino,Other,M,83,UNK,6/9/2022
Non-Hispanic or Latino,Asian,M,79,HET,6/8/2022
Non-Hispanic or Latino,Unknown,M,20,HET,6/11/2022
Hispanic or Latino,Native Hawaiian or Other Pacific Islander,M,61,HET,6/9/2022
Non-Hispanic or Latino,Black or African American,M,32,HET,6/10/2022
Non-Hispanic or Latino,White,F,50,HET,6/14/2022
Hispanic or Latino,Unknown,D,85,HOM,6/13/2022
Non-Hispanic or Latino,Native Hawaiian or Other Pacific Islander,F,4,HET,6/9/2022
Non-Hispanic or Latino,Native Hawaiian or Other Pacific Islander,F,59,HET,6/8/2022
Unknown,Black or African American,M,75,HET,6/9/2022
Non-Hispanic or Latino,Multiple Races,M,45,HET,6/13/2022
Non-Hispanic or Latino,Black or African American,U,56,HET,6/9/2022
Non-Hispanic or Latino,Other,F,36,HET,6/11/2022
Non-Hispanic or Latino,American Indian or Alaska Native,M,79,HET,6/6/2022
Hispanic or Latino,American Indian or Alaska Native,M,88,HET,6/9/2022
Hispanic or Latino,Other,F,82,HET,6/16/2022
Non-Hispanic or Latino,Black or African American,F,7,HET,6/8/2022
Non-Hispanic or Latino,Unknown,D,26,HET,6/11/2022
Non-Hispanic or Latino,Multiple Races,F,75,DNK,6/11/2022
Unknown,Other,D,8,HET,6/8/2022
Non-Hispanic or Latino,Unknown,M,77,HET,6/15/2022
Non-Hispanic or Latino,American Indian or Alaska Native,M,14,HET,6/14/2022
Non-Hispanic or Latino,Other,F,71,HOM,6/8/2022
Non-Hispanic or Latino,Other,F,23,HET,6/10/2022
Non-Hispanic or Latino,White,F,16,HET,6/6/2022
Non-Hispanic or Latino,Asian,F,43,BIS,6/9/2022
Hispanic or Latino,Native Hawaiian or Other Pacific Islander,M,74,HET,6/8/2022
Unknown,White,M,70,HET,6/14/2022
Non-Hispanic or Latino,Multiple Races,M,39,HET,6/10/2022
Non-Hispanic or Latino,Unknown,F,31,HET,6/14/2022
Hispanic or Latino,Asian,M,38,HET,6/9/2022
Non-Hispanic or Latino,Other,F,31,HET,6/8/2022"

header.true <- function(df){
  names(df) <- as.character(unlist(df[1,]))
  df[-1,]
}

linelist <- dplyr::as_tibble(linelist) %>%
  tidyr::separate_rows(value, sep = "\n") %>%
  tidyr::separate_wider_delim(value, delim = ",", names = paste0("V",1:6))
linelist <- header.true(linelist)
linelist$Age <- as.numeric(linelist$Age)
linelist$SpecimenDate <- as.Date(linelist$SpecimenDate, "%m/%d/%Y")
linelist$age_group <- age_groups(linelist$Age, type = "school")
```

Core functions:

-   `theme_apollo` - minimal, clean theme for plots
-   `apollo_label` - labels that match branding aesthetic
-   `end_points` - subset data at last time series data point for labeling line plots
-   `wrap_labels` - wrap long axis labels to better fit plot
-   `geom_dumbbell` - dumbbell-style plot
-   `geom_lollipop` - lollipop-style plot
-   `highlight_geom` - highlight geom\_\* of interest based on expression

## theme_apollo & apollo_label

Basic plot using ggplot2:

```{r boring-plot, fig.height=8, fig.width=10}
linelist %>%
  count(age_group) %>%
  ggplot(aes(x = age_group, y = n)) +
  geom_col() +
    labs(
    title = "Title Goes Here",
    subtitle = "Subtitle Goes Here",
    x = "Age Groups (Years)",
    y = "Percentage (%)"
  )
```

Now we apply `theme_apollo` and `apollo_label`. Note: when using `coord_flip` for horizontal bar charts, set `direction` to "horizontal" within `theme_apollo`.

```{r nice-plot, fig.height=7, fig.width=10}
linelist %>%
  count(age_group) %>%
  mutate(
    percent = add_percent(n),
    label = n_percent(n, percent, reverse = TRUE)
  ) %>%
  ggplot(aes(x = age_group, y = percent, label = label)) +
  geom_col(fill = "#353D42") +
  scale_y_continuous(expand = c(0,0), limits = c(0,55)) +
  theme_apollo(direction = "vertical") +
  labs(
    title = "Title Goes Here",
    subtitle = "Subtitle Goes Here\n",
    x = "Age Groups (Years)",
    y = "Percentage (%)"
  ) +
  apollo_label(vjust = -0.3)
```

## Get line plot label data

In some cases, you may want to add labels at the end of a line plot to increase readability. The simplest way to get the data from the last point in a time series is to put filtered data into `geom_text`:

```{r end-points, fig.height=7, fig.width=10}
ggplot(data = ts, aes(x = date, y = scores, group = site, linetype = site, colour = site)) +
  geom_line(linewidth = 1) +
  scale_x_date(breaks = ts$date, date_labels = "%b\n%Y") +
  scale_y_continuous(expand = c(0,0), limits = c(0,100)) +
  theme_apollo(direction = "vertical") +
  scale_colour_manual(values = c(A = "blue", B = "red")) +
  apollo_label(data = end_points(ts, date), aes(label = site), hjust = 0) +
    labs(
    title = "Title Goes Here",
    subtitle = "Subtitle Goes Here\n",
    x = "Date",
    y = "Percentage (%)"
  )
```

## Wrap long axis labels

For categories with long titles (e.g. race/ethnicity), you may need the option to wrap text to better fit under/next to the axis. To specify amount of wrapping, use `width` argument to adjust (lower width = more wrapping, higher width = less wrapping).

Without wrapping:

```{r no-wrap, fig.height=5, fig.width=8}
ggplot(data = re, aes(x = group, y = score, label = score)) +
  geom_col() +
  theme_apollo(direction = "vertical") +
  scale_y_continuous(expand = c(0,0), limits = c(0,100)) +
  labs(
    title = "Title Goes Here",
    subtitle = "Subtitle Goes Here\n",
    x = "Race/Ethnicity",
    y = "Percentage (%)"
  ) +
  apollo_label(vjust = -0.3)
```

With wrapping:

```{r wrap-label, fig.height=5, fig.height=8}
ggplot(data = re, aes(x = group, y = score, label = score)) +
  geom_col() +
  scale_x_discrete(labels = wrap_labels(width = 15)) +
  theme_apollo(direction = "vertical") +
  scale_y_continuous(expand = c(0,0), limits = c(0,100)) +
    labs(
    title = "Title Goes Here",
    subtitle = "Subtitle Goes Here\n",
    x = "Race/Ethnicity",
    y = "Percentage (%)"
  ) +
  apollo_label(vjust = -0.3)
```

## Dumbbell Plot

Dumbbell plots consist of two points connected by a solid line. Useful in showing changes in pre/post test scores or timelines. Dataset requires two x variables (x and xend), y variable (character/factor). Speciy within `geom_dumbbell`: point colors (colour_x and colour_xend), line thickness (linewidth), line color (colour), and point size (size).

```{r cleveland, fig.height=8, fig.width=8}
ggplot(data = cleveland, aes(x = PRE, xend = POST, y = Country)) +
  geom_dumbbell(colour_x = "red", colour_xend = "blue", size = 5, linewidth = 2, colour = "black") +
  labs(title = "Dumbbell Plot", x = "Pre/Post Scores", y = "Country") +
  theme_apollo(direction = "vertical")
```

## Lollipop Plot

Lollipop Plot is similar to Cleveland Dot Plot, but has only one point with a stem.

```{r lollipop, fig.height=8, fig.width=10}
ggplot(data = cleveland, aes(x = Country, y = POST, color = Country)) +
  geom_lollipop(colour = "red", size = 5, linewidth = 1) +
  labs(title = "Lollipop Plot", x = "Country", y = "Post Score") +
  theme_apollo(direction = "vertical")
```

## Highlight Group of Interest

In plots with multiple categories, you may want to draw attention to a specific group - whether it's the highest/lowest value, a group of interest, etc. `highlight_geom` takes two arguments - an expression (similar to what you'd use in `dplyr::filter`), and a color for highlighting. There are additional arguments for points (`alpha`, `size`) and lines (`linewidth`).

The highlighting does work with facets, but this function should be considered experimental and to be used with caution. Currently works well with `geom_col`/`geom_bar`, `geom_line`, and `geom_point`.

```{r highlight, fig.height=8, fig.width=10, warning=FALSE}
ggplot(data = highlight, aes(x = locations, y = scores, label = scores)) +
  geom_col() +
  facet_wrap(~site) +
  highlight_geom(scores == max(scores), pal = "#9E0059") +
  scale_y_continuous(expand = c(0,0), limits = c(0,100)) +
  theme_apollo(direction = "vertical") +
  labs(
    title = "Showcasing how highlight_geom() works\n",
    caption = "Agency/Unit Name Goes Here."
  ) +
  apollo_label(vjust = -0.3)
```
