---
title: "Linking Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{linking_data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message=FALSE, warning=FALSE, echo=FALSE}
library(OCepi)
```

Another type of repetitive task is creating unique identifiers to join disparate datasets. Often this includes using identifiers such as name, date of birth, and for more specific matching, address and/or phone number.

A few things to be aware of when using the methods below:

-   Names will be automatically converted to upper case. Address is not.
-   Date format in both datasets you want to join must be formatted the same. Inconsistent formatting will result in little to no matching.

## match_id_1

`match_id_1` uses the first four characters of the first and last name, full date of birth, and the first 10 characters of the address. It is recommended to tidy up the address using `clean_address` from this package prior to using `match_id_1`.

```{r match_id_1}
match_id_1("Mickey","Mouse","1955-07-17","1313 Disneyland Dr")
```

## match_id_2

`match_id_2` uses the same variables as `match_id_1` except it uses the full address.

```{r match_id_2}
match_id_2("Mickey","Mouse","1955-07-17","1313 Disneyland Dr")
```

## match_id_3

`match_id_3` uses the first four characters of first and last name, full date of birth, and phone number. Phone number can be quite messy - some contain parentheses, dashes, and/or country codes. One option to tidy up phone number is to use `clean_phone` from this package. Please be aware this function looks for valid, 10 digit U.S. based phone numbers. If you are dealing with a mix of international and U.S. phone numbers, other approaches may be needed.

```{r match_id_3}
match_id_3("Mickey","Mouse","1955-07-17","714-781-4636")
match_id_3("Mickey","Mouse","1955-07-17",clean_phone("714-781-4636"))
```

## match_id_4

`match_id_4` is the most sensitive approach of making unique id's. This id uses the first four characters of the first and last name plus date of birth. When using this id to join data, it is recommended to use other identifiers to validate the matches (e.g. phone number, address).

```{r match_id_4}
match_id_4("Mickey","Mouse","1955-07-17")
```
